### 2024-01-11 pnrs.tex
cleaning up proofs

\begin{theorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
    \label{karh:thm}
    Let $\Sigma$ be a finite alphabet.
    The set of \kl{$\Nat$-rational polynomials} over $\Sigma$
    coincides with $\CoveredPoly[\seqof{X_a}{a \in \Sigma}]$.
\end{theorem}

@@

\begin{fact}
    \label{fact:n-poly-n-poly}
    Let $P \in \Nat[\vec{X}]$. Then, $P$
    is a \kl{$\Nat$-rational polynomial}.
\end{fact}

@@

\begin{lemma}
    The \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
\end{lemma}
\begin{proof}[Proof Idea]
    Let $P$ be a \kl{$\Nat$-rational polynomial}.
    Let $\Sigma$  be the finite alphabet such that
    $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$,
    and $f \colon \Sigma^* \to \Nat$ be the \kl{commutative}
    $\Nat$-rational series computing $P$.
    Because $f$ has polynomial growth, and is a $\Nat$-rational series,
    it is a $\Nat$-polyregular function 
    (\cref{polynomial-rational-polyreg:fact}).

@@

\begin{lemma}
    \label{lem:delta-cool}
    Let $\vec{X}$ be a tuple of indeterminate, and $Y$ be a fresh indeterminate.
    Let $P \in \Rel[\vec{X}Y]$ be in $\CorrectPoly$, and $\alpha Y^p M$ be a
    \kl{maximal monomial} of $P$, with $\alpha > 0$.
    Then, there exists $K \geq 0$
    such that $\Delta_{K} \defined P - \alpha (Y - K)^p M$
    has \kl{positive coefficients} for \kl{maximal monomials}
    in any partial evaluation that satisfies $Y > K$ (if it
    gives a value to $Y$).
\end{lemma}
\begin{proof}
    Let $N$ be greater than the sum of the absolute values of all coefficients
    appearing in $P$, and let $K = 2 N$.

@@

Note that we cannot conclude in the above lemma that the result is in
$\CorrectPoly$, because we have a restriction on the range where the variable
$Y$ can be taken.

\begin{lemma}
    \label{lem:correct-to-n-rat}
    Polynomials in $\CorrectPoly$ are
    \kl{$\Nat$-rational polynomials}.
\end{lemma}
\begin{proof}[Proof Sketch]
    We prove by induction on the number $d$ of variables
    and the ``higman ordering of polynomials" (we look at the multiset
    of tuples of things ...) that for every selection
    $K_1, \dots, K_n$ of lower bounds for the different variables
    (that can be $0$),
    functions that are in $\CorrectPoly$ \emph{over tuples of size
    at least nanania} can be computed by a $\Nat$-rational series
    that outputs $0$ outside the domain and the correct value
    inside.

    For $d = 0$,
    the result is true because
    a constant function is a \kl{$\Nat$-rational polynomial}
    if and only if it is \kl{non-negative}.

    Let us now assume that $d > 0$, and in particular
    there is a variable $Y$.
    Consider $P \in \CorrectPoly[\vec{X} Y]$.
    If $Y$ appears at least one in $P$, then it must appear
    in one of the \kl{maximal monomials} of $P$. Otherwise,
    $P \in \CorrectPoly[\vec{X}]$ and we conclude using the
    induction hypothesis.
    Let $\alpha Y^p M$ be a \kl{maximal monomial} in $P$.
    Using \cref{lem:delta-cool},
    we know that there exists $K$ such that
    $P - \alpha (Y - K)^p M$ almost belongs to $\CorrectPoly$,
    and has a strictly smaller "degree multiset".
    By induction hypothesis,
    we conclude that
    the following function is a $\Nat$-rational series
    (beware that we have to consider the lower bounds to apply
    the induction hypothesis):
    \begin{equation*}
        A \defined
        \begin{cases}
            0 & \text{ if some variable is bounded by its } K_i \\
            0 & \text{ otherwise if  } Y \leq K \\
            P - \alpha(Y - K)^pM & \text{ otherwise }
        \end{cases}
        \quad .
    \end{equation*}

    Because $\alpha \geq 0$, it is an easy check
    that the following function is also a $\Nat$-rational series
    \begin{equation*}
        B \defined
        \begin{cases}
            0 & \text{ if some variable is bounded by its } K_i \\
            0 & \text{ otherwise if } Y \leq K \\
            \alpha(Y - K)^pM & \text{ otherwise }
        \end{cases}
    \end{equation*}

    Remark that by fixing some of the variables
    to values 
    below the bounds $K_i$ (and potentially $K$ for $Y$)
    we obtain a \emph{finite} number of polynomials with strictly fewer
    indeterminate.
    As a consequence, the following function is computable
    as a $\Nat$-rational series:
    \begin{equation*}
        C \defined
        \begin{cases}
            P & \text{ if some variable is bounded by its } K_i \\
            P & \text{ otherwise if } Y \leq K \\
            0 & \text{ otherwise }
        \end{cases}
        \quad .
    \end{equation*}

    We conclude because
    $P = A + B + C$, all of which are $\Nat$-rational series.
\end{proof}

\begin{theorem}
    \label{corrected-version:thm}
    Let $P \in \Rel[\vec{X}]$ be a polynomial.
    The following are equivalent:
    \begin{enumerate}
        \item $P \in \CorrectPoly$,
        \item $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
        \item $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
        \item $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
    \end{enumerate}
\end{theorem}

@@

\begin{lemma}
    \label{decide-correct-poly:lem}
    One can decide if a polynomial is in $\CorrectPoly$.
\end{lemma}
\begin{proof}
    It is easy to check that maximal monomials have positive coefficients.
    Assuming that it is the case, remark that 
    every partial valuation $\nu$
    such that $\restr{P}{\nu}$ has a negative maximal coefficient
    must have values that are bounded by the sum of absolute values of
    coefficients appearing in $P$.
    In particular, one can check all possible choices for $\nu$
    within these bounds.
\end{proof}

@@

\begin{lemma}
    f is commutative $\Nat$ polyregular, then when iterated we obtain
    polynomials in $\CorrectPoly$.
\end{lemma}
\begin{proof}
    by induction on the number of letters 
\end{proof}

>>>>

\begin{faketheorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
    \label{karh:thm}
    Let $\Sigma$ be a finite alphabet.
    The set of \kl{$\Nat$-rational polynomials} over $\Sigma$
    coincides with $\CoveredPoly[\seqof{X_a}{a \in \Sigma}]$.
\end{faketheorem}

@@

\begin{remark}
    \label{inductive-correct-poly:remark}
    Let $P \in \Rel[X,\vec{Y}]$ be a polynomial.
    The following are equivalent:
    \begin{enumerate}
        \item $P \in \CorrectPoly$,
        \item The \kl{maximal monomials} of 
            $P$ have \kl{positive coefficients},
            and 
            for all $n \in \Nat$,
            $\restr{P}{X = n} \in \CorrectPoly$.
    \end{enumerate}
    Furthermore, if $P \in \Rel[]$ is a polynomial without
    indeterminate, then $P \in \CorrectPoly$ if and only
    if its \kl{maximal monomial} (the constant term)
    has a \kl{positive coefficient}.
\end{remark}
\begin{proof}
    By induction on the number of indeterminate,
    leveraging the fact that if there exists at least one indeterminate
    $X$, i.e., that $\vec{X} = X \vec{Y}$, then for all $n \in \Nat$
    and partial function $\nu \colon \vec{Y} \to \Nat$,
    we have 
    \begin{equation*}
        \restr{\left(\restr{P}{X = n}\right)}{\nu} = \restr{P}{ \nu[X = n]}
        \quad
        ,
    \end{equation*}
    where $\nu[X = n]$ is the partial function from
    $\vec{X}$ to $\Nat$ that coincides with $\nu$ on $\vec{Y}$
    and maps $X$ to $n$.
\end{proof}

\begin{example}
    \label{n-poly-n-poly:example}
    Let $P \in \Nat[\vec{X}]$. Then, $P$
    is a \kl{$\Nat$-rational polynomial}.
\end{example}
\begin{proof}
    Let $P \in \Nat[\vec{X}]$, and $\nu \colon \vec{X} \to \Nat$
    be a partial function.
    Notice that $\restr{P}{\nu}$ continues to be
    a polynomial with \kl{positive coefficients}, and in particular
    with 
    \kl{maximal monomials} having \kl{positive coefficients}.
\end{proof}

@@

\begin{lemma}
    \label{n-rat-correct:lem}
    The \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
\end{lemma}
\begin{proof}[Proof Idea]
    Let $P$ be a \kl{$\Nat$-rational polynomial}.
    Let $\Sigma$  be the finite alphabet such that
    $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$,
    and $f \colon \Sigma^* \to \Nat$ be the \kl{commutative}
    $\Nat$-rational series computing $P$.
    Because $f$ has polynomial growth, and is a $\Nat$-rational series,
    it is a $\Nat$-polyregular function 
    (\cref{polynomial-rational-polyreg:fact}).

@@

\begin{lemma}
    \label{lem:delta-cool}
    Let $\vec{X}$ be a tuple of indeterminate, and $Y$ be a fresh indeterminate.
    Let $P \in \Rel[\vec{X}Y]$ be in $\CorrectPoly$, and $\alpha Y^p M$ be a
    \kl{maximal monomial} of $P$, with $\alpha > 0$.
    Then, there exists a computable $K \geq 0$
    such that $\Delta_{K} \defined P - \alpha (Y - K)^p M$
    has \kl{positive coefficients} for \kl{maximal monomials}
    in any partial evaluation that satisfies $Y > K$ (if it
    gives a value to $Y$).
\end{lemma}
\begin{proof}
    Let $N$ be greater than the sum of the absolute values of all coefficients
    appearing in $P$, and let $K = 2 N$.

@@

Note that we cannot conclude in the above lemma that the result is in
$\CorrectPoly$, because we have a restriction on the range where the variable
$Y$ can be taken. The upcoming \cref{lem:correct-to-n-rat} is proven using an
induction on the number $d$ of indeterminate, and the degrees of the
\kl{maximal monomials}. To make the induction formal, let us define
$\Monomials[\vec{X}]$ as the (infinite) set of all possible \kl{monomials} over
$\vec{X}$, ordered by \kl{divisibility}. Remark that this set is isomorphic to
$\Nat^{\vec{X}}$ with the product ordering, which is a
\kl{well-quasi-ordering}.

\AP To a polynomial $P$, we assign the set $\MaximalMonomials(P)$ of its
\kl{maximal monomials}, belonging to $\Pfin(\Monomials)$, which remains a
\kl{well-quasi-ordering} when endowed with the \intro{Hoare ordering} of
subsets, that is, $S_1 \hoareleq S_2$ whenever for all \kl{monomial} $M_1 \in
S_1$, there exists a monomial $M_2 \in S_2$, such that $M_1$ \kl{divides}
$M_2$.

\begin{lemma}
    \label{lem:correct-to-n-rat}
    Let $P$ be a polynomial in $\CorrectPoly$.
    Then, 
    $P$ is
    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
\end{lemma}
\begin{proof}
    We prove the result by induction on $\MaximalMonomials(P)$.

    \textbf{Base case:} If the \kl{maximal monomial} of $P$ is a constant term,
    then because $P \in \CorrectPoly$, $P = n \in \Nat$, which is
    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.

    \textbf{Induction:}
    Assume that $\alpha Y^p M \in \MaximalMonomials(P)$
    for some indeterminate $Y$, $\alpha \neq 0$, and $p \geq 1$.
    Because $P \in \CorrectPoly$, we know that $\alpha > 0$.
    Using \cref{lem:delta-cool},
    we know that there exists a computable $K$ such that
    the polynomial
    $\Diff{K} \defined P - \alpha (Y - K)^p M$
    satisfies that for all partial valuations $\nu \colon Y\vec{X} \to \Nat$
    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
    the \kl{maximal monomials} of $\Diff{K}$ have \kl{positive coefficients}.
    In particular, $\Diff{K}(Y - K)$  belongs to $\CorrectPoly$.
    Furthermore, $\MaximalMonomials(\Diff{K}) = \MaximalMonomials(\Diff{K}(Y - K))
    \hoarele \MaximalMonomials(P)$.
    By induction hypothesis,
    we conclude that
    $\Diff{K}(Y - K)$ is \kl{represented}
    by a \kl{star-free $\Nat$-polyregular function} $g_1$,
    \begin{equation*}
        f_1(Y, \vec{X}) \defined
        \begin{cases}
            0 & \text{ if } Y \leq K \\
            \Diff{K} & \text{ otherwise }
        \end{cases}
        \quad .
    \end{equation*}

    Because $\alpha \geq 0$, it is an easy check
    that the following function is also a $\Nat$-rational series
    \begin{equation*}
        f_2(Y, \vec{X}) \defined
        \begin{cases}
            0 & \text{ if } Y \leq K \\
            \alpha(Y - K)^pM & \text{ otherwise }
        \end{cases}
        \quad .
    \end{equation*}

    Because $P \in \CorrectPoly$, for every $n \leq K$, $\restr{P}{Y = n} \in
    \CorrectPoly$ (\cref{inductive-correct-poly:remark}).
    Furthermore,
    $\MaximalMonomials(\restr{P}{Y = n})
    \hoarele \MaximalMonomials(P)$. Hence, by induction hypothesis,
    the following function is \kl{represented} by a
    \kl{star-free $\Nat$-rational series}:
    \begin{equation*}
        f_3(Y, \vec{X}) \defined
        \begin{cases}
            P(Y, \vec{X}) & \text{ if } Y \leq K \\
            0 & \text{ otherwise }
        \end{cases}
        \quad .
    \end{equation*}
    We conclude by remarking that 
    $P = f_1 + f_2 + f_3$.
\end{proof}

We are now ready to state the corrected and generalised version of
\cref{karh:thm}.

\begin{theorem}
    \label{corrected-version:thm}
    Let $P \in \Rel[\vec{X}]$ be a polynomial.
    The following are equivalent:
    \begin{enumerate}
        \item \label{corrected-1:item} $P \in \CorrectPoly$,
        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
    \end{enumerate}
\end{theorem}
\begin{proof}
    The implications 
    \cref{corrected-4:item} $\implies$
    \cref{corrected-3:item} $\implies$
    \cref{corrected-2:item} are obvious.
    \cref{lem:correct-to-n-rat} proves
    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
    while \cref{n-rat-correct:lem}
    proves 
    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
\end{proof}

@@

Before proving \cref{decidable-n-poly:thm}, let us first examine the
effectiveness of \cref{corrected-version:thm}, namely, provide an algorithm to
decide whether a polynomial $P$ belongs to $\CorrectPoly$. The idea will be to
compute a bound to values taken by the valuations $\nu$ appearing in the
definition of $\CorrectPoly$. As a \emph{mise en jambes}, let us consider
separately the case of one indeterminate.

\begin{lemma}
    \label{decide-correct-univariate:lemma}
    Let $P \in \Rel[X]$, there exists a computable $N \in \Nat$,
    such that the following are equivalent:
    \begin{enumerate}
        \item \label{decide-correct-univariate-1:item}
            $P \in \CorrectPoly$,
        \item \label{decide-correct-univariate-2:item}
            The \kl{maximal monomial}
            of $P$ has a \kl{positive coefficient},
            and $P(n) \in \Nat$
            for all $0 \leq n \leq N_0$.
    \end{enumerate}
\end{lemma}
\begin{proof}
    Let us first remark that the implication 
    \cref{decide-correct-univariate-1:item}
    $\implies$ \cref{decide-correct-univariate-2:item} is trivial for every
    $N \in \Nat$.
    For the converse implication,
    let $P \in \Rel[X]$.
    If $P$ is a constant polynomial, then $P \in \CorrectPoly$
    if and only if $P(0) \in \Nat$, and $N \defined 0$
    suffices.
    Otherwise, $P$ is non-constant, and let $n \geq 1$ be the \kl{degree} of $P$.
    Let $P_1$ and $\alpha > 0$ be such that
    $P \defined P_1 + \alpha X^n$. 
    To prove that $P \in \CorrectPoly$, it suffices to show that
    $P(k) \in \Nat$ for all $k \in \Nat$.
    Let us define $\beta$ as the
    maximal absolute value of coefficients appearing in $P_1$, which is
    assumed to be non-zero,
    and $N_0 \defined \beta + 1 \geq 2$.
    Remark that for $X \geq N_0$,
    \begin{align*}
        P(X) &\geq X^n - \beta \sum_{i = 0}^{n-1} X^i \\
             &= X^n - \beta \frac{X^n - 1}{X - 1} \\
             &= \frac{X^{n+1} - X^n - \beta X^n + \beta}{X - 1} \\
             &= \frac{X^n(X - (\beta + 1)) + \beta}{X - 1} \\
             &\geq 0
    \end{align*}
    This proves that $P(k) \in \Nat$ for all $k \geq N_0$,
    and because $P$ satisfies \cref{decide-correct-univariate-2:item},
    $P(k) \in \Nat$ for all $1 \leq k \leq N_0$. We have proven 
    that $P \in \CorrectPoly$.
\end{proof}

In the case of multiple indeterminate, we leverage
\cref{lem:delta-cool} to perform an induction on the degree
of the polynomials and the number of variables of the polynomial.

\begin{lemma}
    \label{decide-correct-poly:lem}
    One can decide if a polynomial belongs to $\CorrectPoly$.
\end{lemma}
\begin{proof}
    We prove the result by induction on the number of indeterminate.

    \textbf{Base Case:} If $P$ has zero indeterminate, i.e., is a constant
    function, then to decide whether $P \in \CorrectPoly$ it suffices to check
    that $P(0) \in \Nat$.

    \textbf{Induction:} Let $P \in \Rel[Y, \vec{X}]$, we start by checking
    whether \kl{maximal monomials} of $P$ have \kl{positive coefficients}.
    Without loss of generality, assume that $Y$ divides some \kl{maximal
    monomial} $\alpha Y^p M$ with $p \geq 1$. By applying
    \cref{lem:delta-cool}, we can compute a bound $K \in \Nat$ such that
    $\Diff{K} \defined P - \alpha (Y - K)^p M$ has \kl{positive coefficients}
    for its \kl{maximal monomials} for all partial valuations $\nu \colon
    Y\vec{X} \to \Nat$ such that $Y > K$ (if $Y$ is assigned a value). In
    particular, we conclude that, for all $\restr{P}{Y = n}$ belongs to
    $\CorrectPoly$ for all $n > K$. By induction hypothesis, one can
    decide whether $\restr{P}{X = n}$ belongs to $\CorrectPoly$
    for all $0 \leq n \leq K$, which is equivalent 
    to deciding whether
    $P \in \CorrectPoly$ thanks to \cref{inductive-correct-poly:remark}.
\end{proof}

@@
